Data Processing

### 2024-04-02 code/analysis-code/analysis2.qmd
Transferred code from .R to .qmd, updated README
Updated README to reflect changes and transfers

#### Loading library and data set
``` {R}
# Load required libraries
library(MASS) # for robust regression
library(caret) # for cross-validation
library(h2o) #For dimension reduction
library(dplyr)

>>>>

#### Loading library and data set
``` {R, warnings=FALSE, messages=FALSE}
# Load required libraries
library(MASS) # for robust regression
library(caret) # for cross-validation
library(h2o) #For dimension reduction
library(dplyr)

### 2024-04-02 code/eda-code/EDA-code.qmd
Transferred code from .R to .qmd, updated README
Updated README to reflect changes and transfers

---
title: "Exploratory Analysis on CDC Teen Vaccination Surveys"
date: "`r format(Sys.Date()'"
Author: Kelly Cao and Rachel Robertson
output: html_document
editor: 
  markdown: 
    wrap: sentence
---

@@

## Loading the .R script which performed the analysis

```{r, include=FALSE, cache=FALSE}
library(here)
knitr::read_chunk(here("./code/eda-code/edacode.R"))
```

## Loading the Packages

```{r, packages, message = FALSE, warning = FALSE}
```

```{r, loaddata}
```

@@

```{r, table1}

@@

```{r, INCPORAR_I}

@@

```{r, STATE}

@@

```{r, INCQ298A}

@@

```{r, RACEETHK}

@@

```{r, RACE_K}

@@

```{r, INS_STAT2_I}

```{r, FACILITY}

@@

```{r, Vaccination}

@@

We examine the percentage of up-to-date vaccination status by state.
```{r, state-vaccine}

@@

```{r, facility-vaccine}

```{r, race-ethnicity-vaccine}

@@

```{r, fitfig1}

@@

```{r, fitfig2}

@@

### Fit 3
```{r, fitfig3}

### Fit 4
```{r, fitfig4}

>>>>

---
title: "Exploratory Analysis on CDC Teen Vaccination Surveys"
date: 02/21/2024
Author: Kelly Cao and Rachel Robertson
output: html_document
editor: 
  markdown: 
    wrap: sentence
---

@@

## Loading the Packages and Data Set

```{r, warnings=FALSE, message=FALSE}
#load needed packages. make sure they are installed.
library(here) #for data loading/saving
library(dplyr)
library(skimr)
library(ggplot2)
library(knitr)
```

```{r}
#Path to data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","cleandata1.rds")
#load data
mydata <- readRDS(data_location)
```

@@

```{r}
summary_df = skimr::skim(mydata)
print(summary_df)
# save to file
summarytable_file = here("results","tables", "summarytable.rds")
saveRDS(summary_df, file = summarytable_file)
```

@@

```{r}
p1 <- mydata %>% ggplot(aes(x=INCPORAR_I)) + geom_histogram() 
plot(p1)
figure_file = here("results", "figures", "income.png")
ggsave(filename = figure_file, plot=p1) 
```

@@

```{r}
summary(mydata$STATE)
state_distribution <- table(mydata$STATE)
# Convert STATE to character
mydata$STATE <- as.character(mydata$STATE)

# Calculate distribution percentage
state_distribution_percentage <- prop.table(table(mydata$STATE))*100

print(state_distribution_percentage)

p2 <- barplot(state_distribution_percentage, 
        main = "State Distribution Percentage", 
        xlab = "State", 
        ylab = "Percentage",
        col = "skyblue",   # Set bar color
        las = 2) 
#Recreating the graph in ggplot

#adjusting the percent disitrbution data into a data frame
data_df <- as.data.frame(as.table(state_distribution_percentage))

#plot with ggplot
p3 <- ggplot(data_df, aes(x = Var1, y = Freq)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "State Distribution Percentage",
       x = "State",
       y = "Percentage") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

#Viewing the plot to confirm it is done properly
print(p3)

figure_file = here("results", "figures", "state.distribution.png")
ggsave(filename = figure_file, plot=p3)

@@

```{r}
summary(mydata$INCQ298A)
Family_income_distribution <- table(mydata$INCQ298A)

# Calculate distribution percentage
Family_Income_distribution <- prop.table(table(mydata$INCQ298A))*100

print(Family_Income_distribution)

p4 <- barplot(Family_Income_distribution, 
              main = "Family Income Distribution Percentage", 
              xlab = "Income levels", 
              ylab = "Percentage",
              col = "skyblue",   # Set bar color
              las = 2) +
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

#Recreating the graph in ggplot

#adjusting the percent disitrbution data into a data frame
family_income_df <- as.data.frame(as.table(Family_Income_distribution))

#plot with ggplot
p5 <- ggplot(family_income_df, aes(x = Var1, y = Freq)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Family Income Distribution Percentage",
       x = "Income",
       y = "Percentage")

#Viewing the plot to confirm it is done properly
print(p5)

figure_file = here("results", "figures", "family.income.distribution.png")
ggsave(filename = figure_file, plot=p5)

@@

```{r}
summary(mydata$RACEETHK)
ethnicity_distribution <- table(mydata$RACEETHK)

# Calculate distribution percentage
Ethnicity_distribution <- prop.table(table(mydata$RACEETHK))*100

print(Ethnicity_distribution)

p6 <- barplot(Ethnicity_distribution, 
              main = "Ethnicity Distribution Percentage", 
              xlab = "Ethnicity", 
              ylab = "Percentage",
              col = "skyblue",   # Set bar color
              las = 2) +
theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
#Recreating the graph in ggplot

#adjusting the percent disitrbution data into a data frame
ethnicity_df <- as.data.frame(as.table(Ethnicity_distribution))

#plot with ggplot
p7 <- ggplot(ethnicity_df, aes(x = Var1, y = Freq)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Ethnicity Distribution Percentage",
       x = "Ethnicity",
       y = "Percentage")

#Viewing the plot to confirm it is done properly
print(p7)

figure_file = here("results", "figures", "ethnicity.distribution.png")
ggsave(filename = figure_file, plot=p7)

@@

```{r}
summary(mydata$RACE_K)
race_distribution <- table(mydata$RACE_K)

# Calculate distribution percentage
Race_distribution <- prop.table(table(mydata$RACE_K))*100

print(Race_distribution)

p8 <- barplot(Race_distribution, 
              main = "Race Distribution Percentage", 
              xlab = "Race", 
              ylab = "Percentage",
              col = "skyblue",   # Set bar color
              las = 2) 
#Recreating the graph in ggplot

#adjusting the percent disitrbution data into a data frame
race_df <- as.data.frame(as.table(Race_distribution))

#plot with ggplot
p9 <- ggplot(race_df, aes(x = Var1, y = Freq)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Race Distribution Percentage",
       x = "Race",
       y = "Percentage")

#Viewing the plot to confirm it is done properly
print(p9)

figure_file = here("results", "figures", "race.distribution.png")
ggsave(filename = figure_file, plot=p9)
```

@@

```{r}
summary(mydata$INS_STAT2_I)
insurance_status_distribution <- table(mydata$INS_STAT2_I)

# Calculate distribution percentage
Insurance_distribution <- prop.table(table(mydata$INS_STAT2_I))*100

print(Insurance_distribution)

p10 <- barplot(Insurance_distribution, 
              main = "Insurance Status Distribution Percentage", 
              xlab = "Insurance Status", 
              ylab = "Percentage",
              col = "skyblue",   # Set bar color
              las = 2) 
#Recreating the graph in ggplot

#adjusting the percent disitrbution data into a data frame
insurance_df <- as.data.frame(as.table(Insurance_distribution))

#plot with ggplot
p11 <- ggplot(insurance_df, aes(x = Var1, y = Freq)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Insurance Status Distribution Percentage",
       x = "Insurance Status",
       y = "Percentage")

#Viewing the plot to confirm it is done properly
print(p11)

figure_file = here("results", "figures", "insurance.status.distribution.png")
ggsave(filename = figure_file, plot=p11) 
```

```{r}
summary(mydata$FACILITY)
Facility_distribution <- table(mydata$FACILITY)

# Calculate distribution percentage
Facility_distribution <- prop.table(table(mydata$FACILITY))*100

print(Facility_distribution)

p12 <- barplot(Facility_distribution, 
              main = "Facility Distribution Percentage", 
              xlab = "Facility", 
              ylab = "Percentage",
              col = "skyblue",   # Set bar color
              las = 2) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
#Recreating the graph in ggplot

#adjusting the percent disitrbution data into a data frame
facility_df <- as.data.frame(as.table(Facility_distribution))

#plot with ggplot
p13 <- ggplot(facility_df, aes(x = Var1, y = Freq)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Facility Distribution Percentage",
       x = "Facility",
       y = "Percentage") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

#Viewing the plot to confirm it is done properly
print(p13)

figure_file = here("results", "figures", "facility.distribution.png")
ggsave(filename = figure_file, plot=p13)

@@

```{r}
summary(mydata$P_UTDHPV)
Vaccination_status_distribution <- table(mydata$P_UTDHPV)

# Calculate distribution percentage
Vaccination_status_distribution <- prop.table(table(mydata$P_UTDHPV))*100

print(Vaccination_status_distribution)

p14 <- barplot(Vaccination_status_distribution, 
              main = "Vaccination Status Distribution Percentage", 
              xlab = "Vaccination Status", 
              ylab = "Percentage",
              col = "skyblue",   # Set bar color
              las = 2) 
#Recreating the graph in ggplot

#adjusting the percent disitrbution data into a data frame
Vaccine_df <- as.data.frame(as.table(Vaccination_status_distribution))

#plot with ggplot
p15 <- ggplot(Vaccine_df, aes(x = Var1, y = Freq)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Vaccination Status Distribution Percentage",
       x = "Vaccination Status",
       y = "Percentage")

#Viewing the plot to confirm it is done properly
print(p15)

figure_file = here("results", "figures", "vaccination.status.distribution.png")
ggsave(filename = figure_file, plot=p15) 
```

We examine the percentage of up-to-date vaccination status by state.
```{r}
#Aggregate vaccine status stratified by state
vaccine_state_percent <- mydata %>%
  group_by(STATE) %>%
  summarise(percentage_UTD = sum(P_UTDHPV == "UTD") / n() * 100)

#Create table to display only up-to-date status in percents for each state
table1 <- mydata %>%
  filter(P_UTDHPV == "UTD") %>%
  count(STATE) %>%
  rename(UTD_count = n) %>%
  left_join(vaccine_state_percent, by = "STATE") %>%
  select(STATE, percentage_UTD) %>%
  arrange(STATE)

#Print the table
kable(table1, caption = "Vaccine Distribution Percentage per State (Only 'UTD' Status)")

#Write the table to save to the results
write.table(table1, file = here("results","tables", "state-vaccine-completion.txt"), sep = "\t", row.names = FALSE)

@@

```{r}
#Stratifying the HPV vaccine completion by facility
##Make a stacked bar plot to compare which facilities reported UTD patients
p18 <- ggplot(mydata, aes(x = P_UTDHPV, fill = FACILITY)) +
  geom_bar(stat = "count") +  # Use geom_bar() with the count of each facility
  labs(x = "Vaccination Status", y = "Count", title = "Facility Vaccination Completion") +
  theme_minimal()
  print(p18)

#Save the bar plot
figure_file <- here("results", "figures", "facility-vaccination.png")
ggsave(filename = figure_file, plot = p18, width = 8, height = 6, dpi = 300)

@@

```{r}
#Aggregate total count per each race-ethnicity factor level
total_count_by_race <- mydata %>%
  group_by(RACEETHK) %>%
  summarise(total_count = n())

#Aggregate count of those with UTD status for each race-enthnicity factor level
UTD_by_race <- mydata %>%
  filter(P_UTDHPV == "UTD") %>%
  group_by(RACEETHK) %>%
  summarise(UTD_count = n())

#Merge the total count and UTD count data frames
vaccine_percent_by_race <- total_count_by_race %>%
  left_join(UTD_by_race, by = "RACEETHK") %>%
  mutate(percentage_UTD = UTD_count / total_count * 100)

# Print the table
kable(vaccine_percent_by_race, caption = "Percentage of UTD Vaccination Status by RACEETHK")

# Save the table as a text file
write.table(vaccine_percent_by_race, file = here("results", "tables", "vaccine_percent_by_race.txt"), sep = "\t", row.names = FALSE)

@@

```{r}
#Stratifying the Income-poverty levels to the vaccination status in a boxplot figure
p16 <- mydata %>% ggplot(aes(x=P_UTDHPV, y=INCPORAR_I)) + geom_boxplot()
plot(p16)
figure_file = here("results","figures", "income-vaccination.png")
ggsave(filename = figure_file, plot=p16)

# Fitting Income-poverty levels and Vaccination status to a statistical model
model <- lm(INCPORAR_I ~ P_UTDHPV, data = mydata)
summary(model)

@@

```{r}
#Stratifying the Income-poverty levels to the insurance status in a boxplot figure
original_labels <- levels(mydata$INS_STAT2_I)
print(original_labels)

p17 <- mydata %>% 
  ggplot(aes(x=INS_STAT2_I, y=INCPORAR_I, color = P_UTDHPV)) +  
  geom_boxplot() +
  labs(x = "Insurance status", y = "Income-poverty ratio", title = "Income-poverty ratio stratified by insurance and vaccination") +
  scale_x_discrete(labels = c("PRIVATE INSURANCE ONLY" = "Private", "ANY MEDICAID" = "Medicaid", "OTHER INSURANCE (CHIP, IHS, MILITARY, OR OTHER, ALONE OR IN COMB. WITH PRIVATE INSURANCE)" = "Other", "UNINSURED" = "Uninsured")) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))  # Rotate x-axis text by 90 degrees
plot(p17)
figure_file = here("results","figures", "insurance-income-stratified.png")
ggsave(filename = figure_file, plot=p17)

# Fitting Income-poverty levels and insurance status to a statistical model
model <- lm(INCPORAR_I ~ INS_STAT2_I, data = mydata)
summary(model)
```

### Fit 3
```{r}
# Fitting Income-poverty levels and insurance status to a statistical model
model <- lm(INCPORAR_I ~ P_UTDHPV, INS_STAT2_I, data = mydata)
summary(model)
```

### Fit 4
```{r}
#Stratifying the HPV vaccine completion by both Race-Ethnicity and (choose either insurance or income-poverty status)
p19 <- mydata %>% 
  ggplot(aes(x=INCPORAR_I, y=RACEETHK, color = P_UTDHPV)) +  geom_boxplot()
plot(p19)
figure_file = here("results","figures", "raceeth-income-stratified.png")
ggsave(filename = figure_file, plot=p19)

# Fitting Income-poverty levels and insurance status to a statistical model
model <- lm(INCPORAR_I ~ P_UTDHPV, RACEETHK, data = mydata)
summary(model)

@@

### 2024-04-02 code/processing-code/Processing-Quarto-Output.qmd
Transferred code from .R to .qmd, updated README
Updated README to reflect changes and transfers

# Setup

This chunk pulls the R-script entitled Creating_Data_Frame.R.

```{r, include=FALSE, cache=FALSE}
library(here)
knitr::read_chunk(here("./code/processing-code/creating-df/Creating_Data_Frame.R"))
```

Load needed packages.

```{r, packages, message = FALSE, warning = FALSE}
```

# Data loading

```{r, loaddata}
```

# Specifying Factor Variables

```{r, cleandata}

@@

# Save data

```{r, savedata}
```

# Cleaning Data

```{r, include=FALSE, cache=FALSE}
library(here)
knitr::read_chunk(here("./code/processing-code/cleaningdata/cleaning1.R"))
```

```{r, packages2}
```

```{r, loaddata2}
```

## Exploring the data

```{r, exploredata2}
```

## Eliminating NA Values and Columns

```{r, cleandata2}
```

## Saving the Data Frame

```{r, savedata2}
```

>>>>

# Setup
I first load the needed packages.
```{r, message = FALSE, warning = FALSE}
library(readr) #Loading the package to read the fwf
library(here) #Loading the package to produce relative pathway
library(knitr) #Loading in the package for dynamic report generation
library(tidyverse)
library(readxl)
library(dplyr)
```

# Data loading
I load the raw .DAT file into the environment and using the `readr` package, I declare the column position and labels of the desired columns of interest. I then apply the declared object with `read_fwf` function to the raw data (.DAT) file into a data frame named `data_frame`.
```{r}
column_position<- fwf_positions(
  start = c(1, 89, 287, 295, 314, 330, 331, 334, 336, 343, 344, 345,
            347, 353, 359, 362, 472, 473, 495, 496, 497, 502, 504, 1296, 1297
  ),
  end = c(5, 92, 288, 295, 329, 330, 332, 334, 336, 343, 344, 346,
          347, 354, 359, 362, 472, 473, 495, 496, 497, 502, 504, 1296, 1297
  ),
  col_names = c("SEQNUMT", "YEAR", "AGE", "EDUC1", "INCPORAR_I", 
                "INCPOV1", "INCQ298A", "LANGUAGE", "MOBIL_1", "RACEETHK", 
                "RACE_K", "RENT_OWN", "SEX", "STATE", "FACILITY", "WELLCHILD", 
                "P_U13HPV", "P_U13HPV3", "P_UTDHPV", "P_UTDHPV_15", 
                "P_UTDHPV_15INT", "P_UTDHPV2", "P_UTDHPV3", "INS_STAT2_I", 
                "INS_BREAK_I")
)
file_path <- here("data", "raw-data", "NISTEENPUF22.DAT")

data_frame <- read_fwf(file_path, 
                       col_positions = column_position,
                       col_types = cols(.default = "c")
                       )
```

# Specifying Factor Variables
With the data_frame in the environment, I nthen apply the labels for each factor levels using the provided codebook (this codebook can be found in ./data/raw_data/NIS-Teen_Puff22-CODEBOOK.pdf). This will make further downstream analysis much easier to work with.
```{r}
EDUC1levels=c(1,2,3,4)
EDUC1labels=c("LESS THAN 12 YEARS", "12 YEARS", "MORE THAN 12 YEARS, NON-COLLEGE GRAD", "COLLEGE GRADUATE")

INCPOVlevels=c(1,2,3,4)
INCPOVlabels=c("ABOVE POVERTY > $75K", "ABOVE POVERTY <= $75K", "BELOW POVERTY", "UNKNOWN")

INCQ298Alevels=c(10,11,12,13,14,3,4,5,6,7,77,8,9,99)
INCQ298Alabels=c("$35001 - $40000", "$40001 - $50000", "$50001 - $60000", "$60001 - $75000", "$75001+", "$0 - $7500", "$7501 - $10000", "$10001 - $17500", "$17501 - $20000", "$20001 - $25000", "DON'T KNOW", "$25001 - $30000", "$30001 - $35000", "REFUSED")

LANGUAGElevels=c(1,2,3)
LANGUAGElabels=c("ENGLISH", "SPANISH", "OTHER")

MOBILlevels=c(1,2,77,98,99)
MOBILlabels=c("MOVED FROM DIFFERENT STATE", "DID NOT MOVE FROM DIFFERENT STATE", "DON'T KNOW", "MISSING IN ERROR", "REFUSED")

RACEETHKlevels=c(1,2,3,4)
RACEETHKlabels=c("HISPANIC", "NON-HISPANIC WHITE ONLY", "NON-HISPANIC BLACK ONLY", "NON-HISPANIC OTHER + MULTIPLE RACE")

RACE_Klevels=c(1,2,3)
RACE_Klabels=c("WHITE ONLY", "BLACK ONLY", "OTHER + MULTIPLE RACE")

RENTOWNlevels=c(1,2,3,77,99)
RENTOWNlabels=c("OWNED OR BEING BOUGHT", "RENTED", "OTHER ARRANGMENT", "DON'T KNOW", "REFUSED")

SEXlevels=c(1,2,77,98,99)
SEXlabels=c("MALE", "FEMALE", "DON'T KNOW", "MISSING IN ERROR", "REFUSED")

STATElevels=c(1,10,11,12,13,15,16,17,18,19,2,20,21,22,23,24,25,26,27,28,29,30,31,32,33,34,35,36,37,38,39,4,40,41,42,44,45,46,47,48,49,5,50,51,53,54,55,56,6,66,72,78,8,9)
STATElabels=c("ALABAMA", "DELAWARE", "DISTRICT OF COLUMBIA", "FLORIDA", "GEORGIA", "HAWAII", "IDAHO", "ILLINOIS", "INDIANA", "IOWA", "ALASKA", "KANSAS", "KENTUCKY", "LOUISIANA", "MAINE", "MARYLAND", "MASSACHUSETTS", "MICHIGAN", "MINNESOTA", "MISSISSIPPI", "MISSOURI", "MONTANA", "NEBRASKA", "NEVADA",
              "NEW HAMPSHIRE", "NEW JERSEY", "NEW MEXICO", "NEW YORK", "NORTH CAROLINA", "NORTH DAKOTA", "OHIO", "ARIZONA", "OKLAHOMA", "OREGON", "PENNSYLVANIA", "RHODE ISLAND", "SOUTH CAROLINA", "SOUTH DAKOTA", "TENNESSEE", "TEXAS", "UTAH", "ARKANSAS", "VERMONT", "VIRGINIA", "WASHINGTON", "WEST VIRGINIA",
              "WISCONSIN", "WYOMING", "CALIFORNIA", "GUAM", "PUERTO RICO", "U.S. VIRGIN ISLANDS", "COLORADO", "CONNECTICUT")

FACILITYlevels=c(1,2,3,4,5,6, ".")
FACILITYlabels=c("ALL PUBLIC FACILITIES", "ALL HOSPITAL FACILITIES", "ALL PRIVATE FACILITIES", "ALL STD/SCHOOL/TEEN CLINICS OR OTHER FACILITIES", "MIXED", "UNKNOWN", "Missing Data")

WELLCHILDlevels=c(1,2,3)
WELLCHILDlabels=c("YES", "NO", "DON'T KNOW")

UTDlevels=c(0,1, ".")
UTDlabels=c("NOT UTD", "UTD", "Missing Data")

INS_STAT2_Ilevels=c(1,2,3,4, ".")
INS_STAT2_Ilabels=c("PRIVATE INSURANCE ONLY", "ANY MEDICAID", "OTHER INSURANCE (CHIP, IHS, MILITARY, OR OTHER, ALONE OR IN COMB. WITH PRIVATE INSURANCE)", "UNINSURED", "MISSING Data")

INS_BREAK_Ilevels=c(1,2,3,4,".")
INS_BREAK_Ilabels=c("CURRENTLY INSURED BUT UNINSURED AT SOME POINT SINCE AGE 11", "CURRENTLY INSURED AND NEVER UNINSURED SINCE AGE 11", "CURRENTLY UNINSURED BUT INSURED AT SOME POINT SINCE AGE 11", "CURRENTLY UNINSURED AND NEVER INSURED SINCE AGE 11", "Data Missing")

data<-mutate(data_frame,
             SEQNUMT = as.numeric(SEQNUMT), 
             YEAR = as.numeric(YEAR),
             AGE = as.numeric(AGE),
             EDUC1 = factor(EDUC1,levels = EDUC1levels, labels = EDUC1labels),
             INCPORAR_I = as.numeric(INCPORAR_I), 
             INCPOV1 = factor(INCPOV1, levels = INCPOVlevels, labels = INCPOVlabels),
             INCQ298A = factor(INCQ298A, levels = INCQ298Alevels, labels = INCQ298Alabels),             
             LANGUAGE = factor(LANGUAGE, levels = LANGUAGElevels, labels = LANGUAGElabels),
             MOBIL_1 = factor(MOBIL_1, levels = MOBILlevels, labels = MOBILlabels),
             RACEETHK = factor(RACEETHK, levels = RACEETHKlevels, labels = RACEETHKlabels),
             RACE_K = factor(RACE_K, levels = RACE_Klevels, labels = RACE_Klabels),
             RENT_OWN = factor(RENT_OWN, levels = RENTOWNlevels, labels = RENTOWNlabels), 
             SEX = factor(SEX, levels = SEXlevels, labels = SEXlabels),
             STATE = factor(STATE, levels = STATElevels, labels = STATElabels), 
             FACILITY = factor(FACILITY, levels = FACILITYlevels, labels = FACILITYlabels),
             WELLCHILD = factor(WELLCHILD, levels = WELLCHILDlevels, labels = WELLCHILDlabels),
             P_U13HPV = factor(P_U13HPV, levels = UTDlevels, labels = UTDlabels),
             P_U13HPV3 = factor(P_U13HPV3, levels = UTDlevels, labels = UTDlabels),
             P_UTDHPV = factor(P_UTDHPV, levels = UTDlevels, labels = UTDlabels),
             P_UTDHPV_15 = factor(P_UTDHPV_15, levels = UTDlevels, labels = UTDlabels),
             P_UTDHPV_15INT = factor(P_UTDHPV_15INT, levels = UTDlevels, labels = UTDlabels),
             P_UTDHPV2 = factor(P_UTDHPV2, levels = UTDlevels, labels = UTDlabels),
             P_UTDHPV3 = factor(P_UTDHPV3, levels = UTDlevels, labels = UTDlabels),
             INS_STAT2_I = factor(INS_STAT2_I, levels = INS_STAT2_Ilevels, labels = INS_STAT2_Ilabels),
             INS_BREAK_I = factor(INS_BREAK_I, levels = INS_BREAK_Ilevels, labels = INS_BREAK_Ilabels)
)

summary(data)
str(data)

@@

# Save data
We then save the data frame into a .rds and a csv file named "NIS_Teen_Data_2022.csv" and "processeddata.rds"
```{r}
output_file<- here("data", "processed-data", "NIS_Teen_Data_2022.csv")
write.csv(data, file = output_file, row.names = FALSE)
save_path <- here("./data/processed-data/processeddata.rds")
saveRDS(data, file = save_path )
```

# Cleaning Data
```{r, warning=FALSE, message=FALSE}
library(here)
library(dplyr)
library(skimr)
library(ggplot2)
library(tidyr)
library(naniar)
```

First, I load the data into a usable data frame.
```{r}
data_location <- here::here("data","processed-data","processeddata.rds")
NISTeenVax <- readRDS(data_location)
```

## Exploring the data
Next, I will check the structure of the data and I will check for any NA or missing values in the data frame.
```{r}
str(NISTeenVax)
dplyr::glimpse(NISTeenVax)
NA_columns <- colSums(is.na(NISTeenVax))
print(NA_columns)

#There are no missing values, but there may be missing data that is filled with a character value within the data frame. I will first find the character value assigned to missing variables by exploring the levels of each variable using the sapply() and then lapply() function.

factor_cols <- sapply(NISTeenVax, is.factor) #Find the columns that are factor variables
factor_levels <- lapply(NISTeenVax[, factor_cols, drop = FALSE], levels) #Find the levels of each factor variable
print(factor_levels)#Print factor levels for only the factor columns

#Several columns contain factor levels that are labelled, "Data missing", "Missing data", "Unknown", "Refused", Missing in error", and "Don't know". I will replace these values for each variable with an NA in R so that they will be counted as missing values for analysis.
```

## Eliminating NA Values and Columns
Several columns contain factor levels that are labelled, "Data missing", "Missing data", "Unknown", "Refused", Missing in error", and "Don't know". I will replace these values for each variable with an NA in R so that they will be counted as missing values for analysis.
```{r}
missing_values <- c("Data Missing", "MISSING Data", "Missing Data", "MISSING IN ERROR") #Group all of the factor levels that indicate a missing value

#After back and forth with ChatGPT3.5, it suggested to use the dplyr mutate function and replace function to specify that factor variables with the levels in the group above will have those level values replaced with NA
NISTeenVax2 <- NISTeenVax %>% #Create new dataframe
  mutate(across(where(is.factor), ~ replace(., . %in% missing_values, NA))) #mutate only at factor variables in the dataframe and replace the levels specified above with NA for those factors

#Now I will be able to check for NA values using the skim() function. I can also display the NA values graphically using gg_miss_var()

skimr::skim(NISTeenVax2)
gg_miss_var(NISTeenVax2)

#The most important factors for this analysis is the response variables called variations of "P_UTDHPV".
#We have 27,024 missing values for each of these columns. These missing values may have come from the same survey type, however analysis cannot be performed when this crucial data is missing. For this reason, I will find the rows where these missing values take place, ensure that they are all the same rows for the multiple variables, and delete these rows specifically.
#I used chatGPT3,5 to find the function to select for certain rows using dplyr and then filter these for NA values
##I prompted it by using the variable names a,b,c and asked for it to select these columns and filter for the rows with missing variables
rows_with_missing_in_all <- NISTeenVax2 %>%
  select("P_U13HPV", "P_U13HPV3", "P_UTDHPV", "P_UTDHPV_15", "P_UTDHPV_15INT", "P_UTDHPV2", "P_UTDHPV3") %>% #select for columns in question
  filter_all(any_vars(is.na(.))) #Find rows with NA values

##I then prompted ChatGPT3.5 to provide me with code to display the location of these rows so that I may check if they all matched.
if (nrow(rows_with_missing_in_all) > 0) {
  print("Rows with missing values in aforementioned variables")
  print(rows_with_missing_in_all)
} else {
  print("No rows with missing values in aforementioned variables")
}

#It seems that all of the missing rows match for all of these variables. This means that we may delete the rows that contain missing values for all of the specified variables.
#Because the variables listed all contain missing values in the same rows, I will select only one variable to filter.

NISTeenVax3 <- NISTeenVax2 %>%
  filter(!is.na(P_UTDHPV)) #drop the rows with NA values in P_UTDHPV
NISTeenVax3 <- NISTeenVax3 %>%
  filter(!is.na(INS_BREAK_I))
NISTeenVax3 <- NISTeenVax3 %>%
  filter(!is.na(FACILITY))

skimr::skim(NISTeenVax3) #check again for missing values in these variables
gg_miss_var(NISTeenVax3)
skimr::skim(NISTeenVax3) #check again for missing values in these variables
gg_miss_var(NISTeenVax3)

#The majority of data is missing from the WELLCHILD column, which indicates whether the teen had receieved a Well Child visit in recent years. Because this is not crucial to out analysis, I will delete this column. 
#The year column in unnecessary, since all values are from the year 2022. So, I will also delete the YEAR column.

NISTeenVax4 <- NISTeenVax3 %>% #create a new data frame for the filtered data
 select(-WELLCHILD, -YEAR) #Use the select() filter to delete multiple columns from the data frame
str(NISTeenVax4) #Check the remaining variables

## Saving the Data Frame
Save cleaned data as an RDS file.
```{r}
output_file2<- here("data", "processed-data", "NIS_Teen_Data_2022_clean1.csv")
write.csv(NISTeenVax4, file = output_file2, row.names = FALSE)
saved_pathway <- here("./data/processed-data/cleandata1.rds")
saveRDS(NISTeenVax4, file = saved_pathway )
